maxent model avec xvalidation (year-based test training)

### 2019-05-04 src/data_exploration.Rmd
cbi mess, running now for other months

>>>>

```{r}
#jun and oct without gear and with chl_surface in 
vif_nogear_cor_jun <- vif(lm(occurrence ~ temp_depth + salinity_depth + o2_depth + chl_surface + nao_sample + nao_prev + nao_winter + amo_sample + amo_winter, data = junpresab))
write.csv(vif_nogear_cor_jun, "../output/bio/vif/monthly_vifs/vif_nogear_cor_jun_chl.csv", row.names = TRUE)
vif_nogear_cor_jun

vif_nogear_cor_oct <- vif(lm(occurrence ~ temp_depth + salinity_depth + o2_depth + chl_surface + nao_sample + nao_prev + nao_winter + amo_sample + amo_winter, data = junpresab))
write.csv(vif_nogear_cor_jun, "../output/bio/vif/monthly_vifs/vif_nogear_cor_oct_chl.csv", row.names = TRUE)
vif_nogear_cor_jun
```

### 2019-05-04 src/maxent_avec_xvalidation.Rmd
cbi mess, running now for other months

```{r pre-install & load packages, include=FALSE}
options(java.parameters = "-Xmx1g" )#needs to run before dismo loaded - boosts memory for java to 1g
packages <- c("dismo", "stringr", "rJava", "raster", "ecospat", "bossMaps", "lattice", "tidyr")
source("../src/ipak.R")
ipak(packages)
```

@@

```{r}
janvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_winter")
marvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
aprvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
mayvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
junvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
julvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
augvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
sepvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
octvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
novvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
decvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")

@@

select month
```{r month of interest -change}
mthna <- "jan"
mthno <- "1"
```

@@

y1998 = 2.75
y1999 = 2.25
y2000 = 2
y2001 = 1.75
y2002 = 2
y2003 = 2
y2004 = 2
y2005 = 2.25
y2006 = 1.75
y2007 = 1.75
y2008 = 2.25
y2009 = 2.25
y2010 = 2.5
y2011 =  2.5

@@

  if (yr == "2003" | yr == "2004"){
    betaval <- 1
  } else if (yr == "2005"){
    betaval <- 1.5
  } else if (yr == "2006"){
    betaval <- 2.25
  }

    mod <- maxent(x = train, p = trainpa, path = paste0("../output/maxent/models/", timeslice), args=c("maximumiterations=5000", "responsecurves=TRUE", "jackknife=TRUE", "threads=8", "addsamplestobackground=TRUE", paste0("betamultiplier=", betaval)))

@@

      aicetc$train_AUC <- trainauc #training only AUC - not tested
      aicetc$no_trainng_points <- notrainingpoints #numberof points used for training
      aicetc$test_coefficients <- nocoeff #this is to calculate the AIC
      aicetc$test_sum.of.logs <- sumsamplepred #this is to calculate the AIC
      aicetc$test_AUC <- evauc #testing AUC
      aicetc$test_evcor <- evcor #testing correlation coefficient
      aicetc$test_evpcor <- evpcor # testing p-value for correlation coefficient
      aicetc$test_nopresence <- evnp 
      aicetc$test_nobackground <- evna
      aicetc$test_maxsumsenspec <- evmaxtprtnr
      aicetc$test_maxkap <- evmaxkap
      aicetc$test_no_omission <- evno_omission
      aicetc$test_prevalence <- evprevalence
      aicetc$test_equal_sens_spec <- evequal_sens_spec
      aicetc$test_sensitivity <- evsensitivity
      aicetc$test_bic <- bic #not sure this is correct...
      aicetc$beta <- betaval

@@

    unlayer_yr <- subset(unlayer, ascyears == yr)
    unlayer_yr <- unlayer_yr[-c(1), ]
    unlayer_yr[] <- lapply(unlayer_yr, as.character) #from https://stackoverflow.com/questions/2851015/convert-data-frame-columns-from-factors-to-characters
    stackdeplst <- list()
    stackdep <- subset(unlayer_yr, select = c(stackname, ascdepths))
    stackdeplst[[year]] <- stackdep
    write.csv(stackdep, paste0("../output/maxent/predictions/stackdepthlist_", yr, ".csv"), row.names = FALSE)
    aea <- raster("../output/env/aea.tif")

   for (td in 1:nrow(unlayer_yr)){
      temp_depth <- raster(unlayer_yr$temp_depth[td])
      crs(temp_depth) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(temp_depth) <- "temp_depth"
      temp_depth <- projectRaster(temp_depth, aea)
      chl_surface <- raster(unlayer_yr$chl_surface[td])
      crs(chl_surface) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(chl_surface) <- "chl_surface"
      chl_surface <- projectRaster(chl_surface, aea)
      o2_depth <- raster(unlayer_yr$o2_depth[td])
      crs(o2_depth) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(o2_depth) <- "o2_depth"
      o2_depth <- projectRaster(o2_depth, aea)
      salinity_depth <- raster(unlayer_yr$salinity_depth[td])
      crs(salinity_depth) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(salinity_depth) <- "salinity_depth"
      salinity_depth <- projectRaster(salinity_depth, aea)
      nao_sample <- raster(unlayer_yr$nao_sample[td])
      crs(nao_sample) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(nao_sample) <- "nao_sample"
      nao_sample <- projectRaster(nao_sample, aea)
      nao_winter <- raster(unlayer_yr$nao_winter[td])
      crs(nao_winter) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
      names(nao_winter) <- "nao_winter"
      nao_winter <- projectRaster(nao_winter, aea)
      stackn <- unlayer_yr$stackname[td]
      stack <- stack(temp_depth, salinity_depth, o2_depth, chl_surface, nao_sample, nao_winter)
      #stackSave(test,unlayer_yr$stackname[td])
      pred <- predict(mod, stack, file = paste0("../output/maxent/predictions/", mthna, "/", "cloglog_", stackn, ".asc"), overwrite=TRUE, args='outputformat=cloglog') 
      predraw <- predict(mod, stack, file = paste0("../output/maxent/predictions/", mthna, "/", "raw_", stackn, ".asc"), overwrite=TRUE, args=c('outputformat=raw')) 
      prplot <- plot(pred, png(paste0("../output/maxent/predictions/", mthna, "/", "cloglog_", stackn, ".png")))
      dev.off() # stops automatic saving of the plot to a png

    }
}

stdep <- do.call(rbind, stackdeplst)
write.csv(stdep, paste0("../output/maxent/predictions/", mthna, "/stack_depth.csv"), row.names = FALSE)
```

@@

```{r set up threshold binary function}
binmpa <- function(x) {
  ifelse(x <=  mpa_thr, 0,
  ifelse(x >  mpa_thr, 1, NA))
  }
```

@@

for (modtime in 1:nrow(modelmonth)){
  timeslice <- modelmonth[modtime]
  print(timeslice)
  yr <- str_sub(timeslice, -4)
  print(yr)
  thresholds <- read.csv(paste0("../output/maxent/models/", timeslice, "/AICforMaxentReg.csv"), header = TRUE)
  mpa_thr <- thresholds$mpa_0.9_clog
  print(mpa_thr)
  for (cum in 1:length(clogasc_list)){
    cumascyr <- str_sub(clogasc_list[cum], start = 44, end = 47) #selects the year from the file name
    if(yr == cumascyr){
      cumasc <- raster(clogasc_list[cum])
      ascname <- names(cumasc)
      print(ascname)
      cum_mpathr <- calc(cumasc, fun = binmpa)
      writeRaster(cum_mpathr, paste0("../output/maxent/predictions/", mthna,  "/binary_", ascname, ".asc"), overwrite = TRUE)
      plot(cum_mpathr, png(paste0("../output/maxent/predictions/", mthna, "/binary_", ascname, ".png")), overwrite = TRUE)
      dev.off() # stops automatic saving of the plot to a png
    }
  }
}
```

@@

```{r calc no presence cells per layer (avg)} 
ascdepth_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^binary_avg.*asc", full.names = TRUE)
binaryavgstack <- stack(ascdepth_list)
stackstat <- data.frame(matrix(ncol = 2, nrow = nlayers(binaryavgstack)))
stackstat[ ,1] <- c(1:49)
colnames(stackstat) <- c("layer", "cell_sum")

>>>>

```{r pre-install & load packages, include=FALSE}
options(java.parameters = "-Xmx1g" )#needs to run before dismo loaded - boosts memory for java to 1g
packages <- c("dismo", "stringr", "rJava", "raster", "ecospat", "bossMaps", "lattice", "tidyr", "enmSdm")
source("../src/ipak.R")
ipak(packages)
```

@@

```{r}
janvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_winter")
marvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
aprvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
mayvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
junvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chll_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
julvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
augvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
sepvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
octvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
novvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")
decvar <- c("occurrence", "year", "temp_depth", "salinity_depth", "o2_depth", "chl_surface", "nao_sample", "nao_prev", "nao_winter", "amo_sample", "amo_winter")

@@

select month
```{r month of interest -change}
mthna <- "mar"
mthno <- "3"
```

@@

@@

for (year in 1:nrow(modelmonth)){
  timeslice <- modelmonth[year] #testing timeslice
  yr <- str_sub(timeslice, -4) #the year being used for testing
  train <- subset(prback, year != yr) #training data is all years but testing
  trainpa <- train$occurrence #creates a vector of presence/background ID (1/0)
  test <-  subset(prback, year == yr) #creates the testing dataset
  testp <- subset(test, occurrence == "1") #splits the testing dataset into presences
  testp <- subset(testp, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  testb <- subset(test, occurrence == "0") #splits the testing dataset into background
  testb <- subset(testb, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  #mess_pr <- subset(train, occurrence == 1)
  #mess_pr  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling
  train  <- subset(train, select = -c(year, occurrence)) #don't need year and occurrence for modelling

  if (yr == "2003"){
    betaval <- 1
  } else if (yr == "1999" | yr == "2002" | yr == "2008" | yr == "2009") {
    betaval <- 1.25
  } else if (yr == "2005" | yr == "2006"){
    betaval <- 1.5
  } else if (yr == "2000"){
    betaval <- 1.75
  } else if (yr == "2001" | yr == "2007"){
    betaval <- 2.25
  } else if (yr == "2010"){
    betaval <- 3.5
  }

    mod <- maxent(x = train, p = trainpa, path = paste0("../output/maxent/models/", timeslice), args=c("writebackgroundpredictions=TRUE",  "maximumiterations=5000", "responsecurves=TRUE", "jackknife=TRUE", "threads=8", "addsamplestobackground=TRUE", paste0("betamultiplier=", betaval)))

@@

    #continuous boyce index
    predPres <- read.csv(paste0("../output/maxent/models/", timeslice, "/species_samplePredictions.csv"), header = TRUE)
    predBg <- read.csv(paste0("../output/maxent/models/", timeslice, "/species_backgroundPredictions.csv"), header = TRUE)
    predPres <- predPres$Cloglog.prediction
    predBg <- predBg$Cloglog
    cbi <- contBoyce(pres = predPres, bg = predBg)

      aicetc$train_AUC <- trainauc #training only AUC - not tested
      aicetc$no_trainng_points <- notrainingpoints #numberof points used for training
      aicetc$test_coefficients <- nocoeff #this is to calculate the AIC
      aicetc$test_sum.of.logs <- sumsamplepred #this is to calculate the AIC
      aicetc$test_AUC <- evauc #testing AUC
      aicetc$test_evcor <- evcor #testing correlation coefficient
      aicetc$test_evpcor <- evpcor # testing p-value for correlation coefficient
      aicetc$test_nopresence <- evnp 
      aicetc$test_nobackground <- evna
      aicetc$test_maxsumsenspec <- evmaxtprtnr
      aicetc$test_maxkap <- evmaxkap
      aicetc$test_no_omission <- evno_omission
      aicetc$test_prevalence <- evprevalence
      aicetc$test_equal_sens_spec <- evequal_sens_spec
      aicetc$test_sensitivity <- evsensitivity
      aicetc$tss <- max(ev@TPR + ev@TNR) - 1
      aicetc$test_bic <- bic #not sure this is correct...
      aicetc$beta <- betaval
      aicetc$cbi <- cbi

@@

    unlayer_yr <- subset(unlayer, ascyears == yr)
    unlayer_yr <- unlayer_yr[-c(1), ]
    unlayer_yr[] <- lapply(unlayer_yr, as.character) #from https://stackoverflow.com/questions/2851015/convert-data-frame-columns-from-factors-to-characters
    stackdeplst <- list()
    stackdep <- subset(unlayer_yr, select = c(stackname, ascdepths))
    stackdeplst[[year]] <- stackdep
    write.csv(stackdep, paste0("../output/maxent/models/", timeslice, "/stackdepthlist_", yr, ".csv"), row.names = FALSE)
    aea <- raster("../output/env/aea.tif")

    for (td in 1:nrow(unlayer_yr)){
      temp_depth <- raster(unlayer_yr$temp_depth[td])
      crs(temp_depth) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(temp_depth) <- "temp_depth"
      temp_depth <- projectRaster(temp_depth, aea)
      chl_surface <- raster(unlayer_yr$chl_surface[td])
      crs(chl_surface) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(chl_surface) <- "chl_surface"
      chl_surface <- projectRaster(chl_surface, aea)
      o2_depth <- raster(unlayer_yr$o2_depth[td])
      crs(o2_depth) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(o2_depth) <- "o2_depth"
      o2_depth <- projectRaster(o2_depth, aea)
      salinity_depth <- raster(unlayer_yr$salinity_depth[td])
      crs(salinity_depth) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(salinity_depth) <- "salinity_depth"
      salinity_depth <- projectRaster(salinity_depth, aea)
      nao_sample <- raster(unlayer_yr$nao_sample[td])
      crs(nao_sample) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(nao_sample) <- "nao_sample"
      nao_sample <- projectRaster(nao_sample, aea)
      nao_prev <- raster(unlayer_yr$nao_prev[td])
      crs(nao_prev) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(nao_prev) <- "nao_prev"
      nao_prev <- projectRaster(nao_prev, aea)
      nao_winter <- raster(unlayer_yr$nao_winter[td])
      crs(nao_winter) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
      names(nao_winter) <- "nao_winter"
      nao_winter <- projectRaster(nao_winter, aea)
      amo_sample <- raster(unlayer_yr$amo_sample[td])
      crs(amo_sample) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0" 
      names(amo_sample) <- "amo_sample"
      amo_sample <- projectRaster(amo_sample, aea)
      amo_winter <- raster(unlayer_yr$amo_winter[td])
      crs(amo_winter) <- "+proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0"
      names(amo_winter) <- "amo_winter"
      amo_winter <- projectRaster(amo_winter, aea)
      stackn <- unlayer_yr$stackname[td]
      stack <- stack(temp_depth, salinity_depth, o2_depth, chl_surface, nao_sample, nao_prev, nao_winter, amo_sample, amo_winter)

      #stackSave(test,unlayer_yr$stackname[td])
      pred <- predict(mod, stack, file = paste0("../output/maxent/predictions/", mthna, "/", "cloglog_", stackn, ".asc"), overwrite=TRUE, args=c('outputformat=cloglog')) 
      predraw <- predict(mod, stack, file = paste0("../output/maxent/predictions/", mthna, "/", "raw_", stackn, ".asc"), overwrite=TRUE, args=c('outputformat=raw')) 
      prplot <- plot(pred, png(paste0("../output/maxent/predictions/", mthna, "/", "cloglog_", stackn, ".png")))
      dev.off() # stops automatic saving of the plot to a png

      #mess

      #messev <- dismo::mess(x = stack, v = mess_pr, full = FALSE)
      #messplot <- plot(messev, png(paste0("../output/maxent/predictions/", mthna, "/", "mess_", stackn, ".png")))
      #dev.off()
      # not working how i'd like....
    }
}

stdep <- do.call(rbind, stackdeplst)
write.csv(stdep, paste0("../output/maxent/models/", timeslice, "/stack_depth.csv"), row.names = FALSE)
```

@@

```{r set up threshold binary function}
binmpa <- function(x) {
  ifelse(x <  mpa_thr, 0,
  ifelse(x >=  mpa_thr, 1, NA))
  }
```

@@

for (modtime in 1:nrow(modelmonth)){
  timeslice <- modelmonth[modtime]
  print(timeslice)
  yr <- str_sub(timeslice, -4)
  thresholds <- read.csv(paste0("../output/maxent/models/", timeslice, "/AICforMaxentReg.csv"), header = TRUE)
  mpa_thr <- thresholds$mpa_0.9_clog
  for (cum in 1:length(clogasc_list)){
    cumascyr <- str_sub(clogasc_list[cum], start = 44, end = 47) #selects the year from the file name
    if(yr == cumascyr){
      cumasc <- raster(clogasc_list[cum])
      ascname <- names(cumasc)
      cum_mpathr <- calc(cumasc, fun = binmpa)
      writeRaster(cum_mpathr, paste0("../output/maxent/predictions/", mthna,  "/binary_", ascname, ".asc"), overwrite = TRUE)
      plot(cum_mpathr, png(paste0("../output/maxent/predictions/", mthna, "/binary_", ascname, ".png")), overwrite = TRUE)
      dev.off() # stops automatic saving of the plot to a png
    }
  }
}
```

@@

```{r calc no presence cells per layer (avg)} 
ascdepth_list <- list.files(paste0("../output/maxent/predictions/", mthna), pattern = "^binary_avg_cloglog.*asc", full.names = TRUE)
binaryavgstack <- stack(ascdepth_list)
stackstat <- data.frame(matrix(ncol = 2, nrow = nlayers(binaryavgstack)))
stackstat[ ,1] <- c(1:49)
colnames(stackstat) <- c("layer", "cell_sum")
